---
title: "Bayesian Assignment 1"
author: "Aphinda Jabe"
format: pdf
editor: visual
---

## Quarto

```{r}
rm(list = ls())
# Packages required
require ( MASS )
library(tidyverse)
```

You can add options to executable code like this

```{r}
#| echo: false

# Load the Auto MPG dataset
auto <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data",
  header = FALSE,
  sep = "",
  col.names = c("mpg", "cylinders", "displacement", "horsepower",
                "weight", "acceleration", "model_year", "origin", "car_name"),
  na.strings = "?"
)

# Remove missing values
auto <- na.omit(auto)

# Response variable
Y <- matrix(auto$mpg, ncol = 1)

# Predictor variables
Xpreds <- auto[, c("displacement", "horsepower", "weight", "acceleration")]

# Design matrix (add intercept + standardise predictors)
X <- as.matrix(cbind(1, scale(Xpreds)))

# Sample size
n <- nrow(X)

```

# Gibbs Sampling

```{r}
gibbs <- function(y,x,n_iter){ 
  k = ncol(x)
  
  # Priors
  beta_til = rep(0,k)
  M = diag(k)
  a = 1
  b = 1
  
  beta_samples = matrix(0, n_iter, k)
  var_samples = numeric(n_iter)
  
  #Starting values
  beta = rep(0, k)
  var = 1
  
  for (i in 1:n_iter){ 
    mu_beta <- solve(M + t(x)%*%x)%*%(t(x)%*%y + M%*%beta_til)
    
    # Sampling the betas
    beta <- MASS::mvrnorm(1, mu = mu_beta, Sigma = var * solve(M + t(x)%*%x))
                          
    #Sample from sigmasqrd
    A2 <- t(y)%*%y + t(beta_til) %*% M %*% beta_til - t(mu_beta) %*% (M + t(x) %*%x) %*% mu_beta
    
    shape = a + n/2
    rate = b + A2/2
    
    var = 1/rgamma(1, shape = shape, rate = rate)
    
    beta_samples[i,] = beta
    var_samples[i]  = var
  }
    return(list(beta = beta_samples, sigma2 = var_samples))
    
  
  }

par_data = gibbs(Y, X, 50000)
```

```{r}
samples <- par_data

# Trace plots for beta
par(mfrow = c(2, 2))  

for (j in 1:ncol(samples$beta)) {
  plot(samples$beta[, j], type = "l",
       main = paste("Trace plot for beta", j),
       ylab = "", xlab = "Iteration")
}
```

The trace plots fluctuate around a stable mean and do not show any visible trends or drift over iterations. This suggests that the Markov chains have reached a stationary distribution. Therefore, we can conclude that the Gibbs sampler appears to have converged.

```{r}
# OLS estimates
ols <- solve(t(X) %*% X) %*% t(X) %*% Y

# Posterior means
post_means <- colMeans(par_data$beta)

par(mfrow = c(2,3))  # adjust depending on number of betas

for (j in 1:ncol(par_data$beta)) {
  
  dens <- density(par_data$beta[,j])
  plot(dens, main = paste("Beta", j), xlab = "")
  
  # Posterior mean
  abline(v = post_means[j], lty = 1, col = "red")
  
  # OLS estimate
  abline(v = ols[j], lty = 2, col = "blue")
}
```

The posterior densities are approximately symmetric and unimodal. The posterior means lie close to the OLS estimates.

```{r}
par(mfrow = c(2,3))

for (j in 1:ncol(par_data$beta)) {
  
  dens <- density(par_data$beta[,j])
  plot(dens, main = paste("Beta", j), xlab = "")
  
  # Credible interval
  ci <- quantile(par_data$beta[,j], c(0.025, 0.975))
  abline(v = ci, lty = 1)
  
  # OLS confidence interval
  residuals <- Y - X %*% ols
  sigma2_hat <- sum(residuals^2) / (nrow(X) - ncol(X))
  var_beta <- sigma2_hat * solve(t(X) %*% X)
  se <- sqrt(diag(var_beta))
  
  ci_ols <- c(ols[j] - 1.96*se[j], ols[j] + 1.96*se[j])
  abline(v = ci_ols, lty = 2)
}
```

```{r}
#### Required Functions ####

library(ggplot2)
library(MASS)

#### Data Generating Functions ####

generate_lost <- function(grid_size, nsims){
  
  # Function to generate the prior distribution for the 
  # location of the lost fisherman
  # Args: 
  #       grid_size: the dimensions of the square search grid
  #       nsims: number of samples to base the prior distribution on
  
  mu_vec  <- c(grid_size/2, grid_size/2)
  sig_mat <- matrix(c(2, 1, 5, 5), 2,2)
  
  dat <- mvrnorm(nsims, mu_vec, sig_mat)
  dat <- round(abs(dat))
  
  prior <- matrix(rep(0,grid_size^2), grid_size, grid_size)
  for (i in 1:NROW(dat)){
    
    if (dat[i,1] < grid_size & dat[i,2] < grid_size){
      prior[dat[i,1], dat[i,2]] <- prior[dat[i,1], dat[i,2]] + 1
    }
    
  }
  prior <- prior/sum(prior)
  return(prior)
}

generate_fisherman <- function(grid_size){
  
  # Function to generate the true location of the lost fisherman.
  # This should not effect the search decision in any way!! It is unkown
  # to the search crew.
  # Args: 
  #       grid_size: the dimensions of the square search grid

  
  mu_vec  <- c(grid_size/2, grid_size/2)
  sig_mat <- matrix(c(2, 1, 5, 5), 2,2)
  
  location  <- round(mvrnorm(1, mu_vec, sig_mat))
  true_grid <- matrix(rep(0, grid_size^2), grid_size, grid_size)
  true_grid[location[1], location[2]] <- 1
  
  return(true_grid)
}

###Question b abd c
grid_size <- 20
nsims <- 10000

theta <- generate_lost(grid_size, nsims)
true_grid <- generate_fisherman(grid_size)

# Detection probabilities (example: uniform)
p <- matrix(0.6, grid_size, grid_size)

time <- 0
found <- FALSE

theta_history <- list()
true_prob_track <- c()

true_loc <- which(true_grid == 1, arr.ind = TRUE)

while(!found){
  
  time <- time + 1
  
  # Step 1: compute detection probability
  detect_prob <- theta * p
  
  # Step 2: choose best cell
  idx <- which(detect_prob == max(detect_prob), arr.ind = TRUE)[1,]
  i <- idx[1]
  j <- idx[2]
  
  # Step 3: simulate detection
  if(true_grid[i,j] == 1){
    found <- rbinom(1,1,p[i,j]) == 1
  } else {
    found <- FALSE
  }
  
  # Record probability of true cell
  true_prob_track <- c(true_prob_track, theta[true_loc[1], true_loc[2]])
  
  # Save heatmap states (first 4 steps)
  if(time %in% c(1,5,10,20)){
    theta_history[[length(theta_history)+1]] <- theta
  }
  
  # Step 4: update if not found
  if(!found){
    
    denom <- 1 - p[i,j]*theta[i,j]
    
    # Update searched cell
    theta[i,j] <- ((1 - p[i,j]) * theta[i,j]) / denom
    
    # Update all other cells
    for(a in 1:grid_size){
      for(b in 1:grid_size){
        if(!(a == i & b == j)){
          theta[a,b] <- theta[a,b] / denom
        }
      }
    }
    
  }
}

#========
#  i
#=========
plot_heatmap <- function(mat, title){
  df <- reshape2::melt(mat)
  ggplot(df, aes(Var1, Var2, fill=value)) +
    geom_tile() +
    scale_fill_viridis_c() +
    ggtitle(title) +
    theme_minimal()
}

plot_heatmap(theta_history[[1]], "Step 1")
plot_heatmap(theta_history[[2]], "Step 5")
plot_heatmap(theta_history[[3]], "Step 10")
plot_heatmap(theta_history[[4]], "Step 20")

#=====
#ii
#======

plot(true_prob_track, type="l",
     xlab="Search Step",
     ylab="Posterior Probability",
     main="Probability of True Cell Over Time")

```
### Posterior conditional distributions (a)

#### First derivation

$$
\begin{align*}p(\beta \mid \tau_1, \tau_2, X, Y) &\propto p(Y \mid X, \beta, \tau_1, \tau_2) \, p(\beta) \\\\\quad p(Y \mid X, \beta, \tau_1, \tau_2) &\propto \exp\left\{ -\frac{1}{2} \left[ \tau_1 (Y_1 - X_1\beta)^T (Y_1 - X_1\beta) + \tau_2 (Y_2 - X_2\beta)^T (Y_2 - X_2\beta)\right] \right\} \\\\\quad p(\beta \mid \tau_1, \tau_2, X, Y) &\propto \exp\left\{ -\frac{1}{2} \left[ \beta^T T_0^{-1} \beta + \tau_1 (Y_1 - X_1\beta)^T (Y_1 - X_1\beta) + \tau_2 (Y_2 - X_2\beta)^T (Y_2 - X_2\beta)\right] \right \} \\\\\quad &\propto \exp\left\{ -\frac{1}{2} \left[\beta^T \left( \tau_1 X_1^T X_1 + \tau_2 X_2^T X_2 + T_0^{-1} \right) \beta - 2\beta^T \left( \tau_1 X_1^T Y_1 + \tau_2 X_2^T Y_2 \right)\right] \right\} \\\\\quad \beta \mid \tau_1, \tau_2, X, Y &\sim N(\mu_n, \Sigma_n) \\\quad \Sigma_n^{-1} &= \tau_1 X_1^T X_1 + \tau_2 X_2^T X_2 + T_0^{-1} \\\mu_n &= \Sigma_n \left( \tau_1 X_1^T Y_1 + \tau_2 X_2^T Y_2 \right)\end{align*}
$$

#### Second derivation

$$
\begin{align*}p(\tau_1 \mid \beta, \tau_2, X, Y) &\propto p(Y_1 \mid X_1, \beta, \tau_1) \, p(\tau_1) \\\\\quad p(\tau_1) &\propto \tau_1^{a-1} \exp(-b\,\tau_1) \\\\\quad p(Y_1 \mid X_1, \beta, \tau_1) &\propto \tau_1^{n_1/2} \exp\left\{ -\frac{\tau_1}{2} (Y_1 - X_1\beta)^T (Y_1 - X_1\beta) \right\} \\ \\\\\quad p(\tau_1 \mid \beta, \tau_2, X, Y) &\propto \tau_1^{a-1} \exp(-b\,\tau_1) \times \tau_1^{n_1/2} \exp\left\{ -\frac{\tau_1}{2} \sum_{i=1}^{n_1} (y_{1i} - x_{1i}^T\beta)^2 \right\} \\&\propto \tau_1^{a + \frac{n_{1}}{2} - 1} \exp\left\{ -\tau_1 \left[ b + \frac{1}{2} \sum_{i=1}^{n_1} (y_{1i} - x_{1i}^T\beta)^2 \right] \right\} \\\\\quad \tau_1 \mid \beta, \tau_2, X, Y &\sim \left( a + \frac{n_1}{2}, \; b + \frac{1}{2} \sum_{i=1}^{n_1} (y_{1i} - x_{1i}^T\beta)^2 \right)\end{align*}
$$

#### Third derivation

\begin{align*}p(\tau_2 \mid \tau_1, \beta, X, Y)&\propto p(Y_2 \mid X_2, \beta, \tau_2) \, p(\tau_2 \mid \tau_1) \\\\(\tau_2 \mid \tau_1) &\propto \tau_2^{a-1} \exp(-b\,\tau_2) \, \mathbb{I}(0 < \tau_2 < \tau_1) \\\\(Y_2 \mid X_2, \beta, \tau_2)&\propto \tau_2^{n_2/2} \exp\left\{ -\frac{\tau_2}{2} (Y_2 - X_2\beta)^T (Y_2 - X_2\beta) \right\} \\\\(\tau_2 \mid \tau_1, \beta, X, Y)&\propto \tau_2^{a + n_2/2 - 1} \exp\left\{ -\tau_2 \left[ b + \frac{1}{2} (Y_2 - X_2\beta)^T (Y_2 - X_2\beta) \right] \right\} \mathbb{I}(0 < \tau_2 < \tau_1) \\\\\tau_2 \mid \tau_1, \beta, X, Y &\sim \left( a + \frac{n_2}{2}, \; b + \frac{1}{2} \sum_{i=1}^{n_2} (y_{2i} - x_{2i}^T\beta)^2 \right) \mathbb{I}(0 < \tau_2 < \tau_1)\end{align*}
$$

### Sampling (b)

$$
\begin{align*}p(\beta, \tau_1, \tau_2 \mid X, Y)&\propto p(Y \mid X, \beta, \tau_1, \tau_2) \, p(\beta) \, p(\tau_1, \tau_2) \\\\(Y \mid X, \beta, \tau_1, \tau_2)&\propto \tau_1^{n_1/2} \tau_2^{(n-n_1)/2}\exp\left\{ -\frac{1}{2} \left[\tau_1 (Y_1 - X_1\beta)^T (Y_1 - X_1\beta)+ \tau_2 (Y_2 - X_2\beta)^T (Y_2 - X_2\beta)\right] \right\} \\\\p(\beta) &\propto \exp\left\{ -\frac{1}{2} \beta^T T_0^{-1} \beta \right\} \\\\p(\tau_1, \tau_2) &\propto \tau_1^{a-1} \tau_2^{a-1} \exp\left\{ -b(\tau_1 + \tau_2) \right\} \mathbb{I}(0 < \tau_2 < \tau_1) \\\\(\beta, \tau_1, \tau_2 \mid X, Y)&\propto \tau_1^{a + n_1/2 - 1} \tau_2^{a + (n-n_1)/2 - 1} \\&\quad \times \exp\left\{ -\frac{1}{2}\beta^T T_0^{-1} \beta- \frac{\tau_1}{2} (Y_1 - X_1\beta)^T (Y_1 - X_1\beta)- \frac{\tau_2}{2} (Y_2 - X_2\beta)^T (Y_2 - X_2\beta)- b(\tau_1 + \tau_2) \right\} \\&\quad \times \mathbb{I}(0 < \tau_2 < \tau_1)\end{align*}
$$

```{r, fig.show='hide'}
# Packages required
require(MASS)
# Load the Auto MPG dataset
auto <-read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data",
  header = FALSE, sep = "",
  col.names = c( "mpg" , "cylinders" , "displacement" , "horsepower" ,
                 "weight" , "acceleration" , "model_year" , "origin" , "car_name" ),
  na.strings = "?"
)
auto <-na.omit(auto)
Y <-matrix(auto$mpg, ncol = 1)
Xpreds <-auto[, c( "displacement" , "horsepower" , "weight" , "acceleration" )]
X <-as.matrix(cbind(1, scale(Xpreds)))
n <-nrow(X)

#### Required Functions ####

library(ggplot2)
library(MASS)

# Contaminate the data with outliers
set.seed(12)
n1 <-350 # Number of 'standard' observations
n2 <-n- n1 # Number of outliers
outlier.idx <-(n1 + 1):n # Index set I2
# Inflate the response for the outlier observations
Y.contaminated <-Y
Y.contaminated[outlier.idx] <-Y[outlier.idx] +
  rnorm(n2, mean = 0, sd = 15)

X1 <- X[1:n1,]
X2 <- X[(n1+1):n,]
y1 <- Y.contaminated[1:n1,]
y2 <- Y.contaminated[outlier.idx]

a1 <- 0.1
b1 <- 0.1
a2 <- 0.1
b2 <- 0.1
T_0 <- diag(0.001, ncol(X1))

iterations <- 70000
betas <- matrix(0, iterations, ncol(X1))
tau1s <- tau2s <- numeric(iterations)

tau1 <- 1
tau2 <- 0.5

for (i in 1:iterations) {
  
  post_variance <- solve(tau1 * crossprod(X1) + tau2 * crossprod(X2) + solve(T_0))
  mu <- post_variance %*% (tau1 * t(X1) %*% y1 + tau2 * t(X2) %*% y2)
  beta <- mvrnorm(1, mu, post_variance)
  
  rss1 <- sum((y1 - X1 %*% beta)^2)
  tau1 <- rgamma(1, a1 + n1/2, b1 + 0.5*rss1)
  
  rss2 <- sum((y2 - X2 %*% beta)^2)
  repeat {tau2_temp <- rgamma(1, shape = a2 + n2/2, rate = b2 + 0.5*rss2)
  if (tau2_temp <- tau1) break}
  tau2 <- tau2_temp
  
  betas[i,] <- beta
  tau1s[i] <- tau1
  tau2s[i] <- tau2
  
}

betas <- betas[(0.2 * iterations + 1):iterations,]
tau1s <- tau1s[(0.2 * iterations + 1):iterations]
tau2s <- tau2s[(0.2 * iterations + 1):iterations]

par(mfrow = c(1, 3))
plot(betas[,1], type = "l")
plot(tau1s, type = "l")
plot(tau2s, type = "l")
```

### Trace plots (c(i))

![](plot){fig-align="center" width="100%"}
