---
title: "Bayesian Assignment 1"
author: "Aphinda Jabe"
format: pdf
editor: visual
---

## Quarto

```{r}
rm(list = ls())
# Packages required
require ( MASS )
library(tidyverse)
```

You can add options to executable code like this

```{r}
#| echo: false

# Load the Auto MPG dataset
auto <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data",
  header = FALSE,
  sep = "",
  col.names = c("mpg", "cylinders", "displacement", "horsepower",
                "weight", "acceleration", "model_year", "origin", "car_name"),
  na.strings = "?"
)

# Remove missing values
auto <- na.omit(auto)

# Response variable
Y <- matrix(auto$mpg, ncol = 1)

# Predictor variables
Xpreds <- auto[, c("displacement", "horsepower", "weight", "acceleration")]

# Design matrix (add intercept + standardise predictors)
X <- as.matrix(cbind(1, scale(Xpreds)))

# Sample size
n <- nrow(X)

```

# Gibbs Sampling

```{r}
gibbs <- function(y,x,n_iter){ 
  k = ncol(x)
  
  # Priors
  beta_til = rep(0,k)
  M = diag(k)
  a = 1
  b = 1
  
  beta_samples = matrix(0, n_iter, k)
  var_samples = numeric(n_iter)
  
  #Starting values
  beta = rep(0, k)
  var = 1
  
  for (i in 1:n_iter){ 
    mu_beta <- solve(M + t(x)%*%x)%*%(t(x)%*%y + M%*%beta_til)
    
    # Sampling the betas
    beta <- MASS::mvrnorm(1, mu = mu_beta, Sigma = var * solve(M + t(x)%*%x))
                          
    #Sample from sigmasqrd
    A2 <- t(y)%*%y + t(beta_til) %*% M %*% beta_til - t(mu_beta) %*% (M + t(x) %*%x) %*% mu_beta
    
    shape = a + n/2
    rate = b + A2/2
    
    var = 1/rgamma(1, shape = shape, rate = rate)
    
    beta_samples[i,] = beta
    var_samples[i]  = var
  }
    return(list(beta = beta_samples, sigma2 = var_samples))
    
  
  }

par_data = gibbs(Y, X, 50000)
```

```{r}
samples <- par_data

# Trace plots for beta
par(mfrow = c(2, 2))  

for (j in 1:ncol(samples$beta)) {
  plot(samples$beta[, j], type = "l",
       main = paste("Trace plot for beta", j),
       ylab = "", xlab = "Iteration")
}
```

The trace plots fluctuate around a stable mean and do not show any visible trends or drift over iterations. This suggests that the Markov chains have reached a stationary distribution. Therefore, we can conclude that the Gibbs sampler appears to have converged.

```{r}
# OLS estimates
ols <- solve(t(X) %*% X) %*% t(X) %*% Y

# Posterior means
post_means <- colMeans(par_data$beta)

par(mfrow = c(2,3))  # adjust depending on number of betas

for (j in 1:ncol(par_data$beta)) {
  
  dens <- density(par_data$beta[,j])
  plot(dens, main = paste("Beta", j), xlab = "")
  
  # Posterior mean
  abline(v = post_means[j], lty = 1, col = "red")
  
  # OLS estimate
  abline(v = ols[j], lty = 2, col = "blue")
}
```

The posterior densities are approximately symmetric and unimodal. The posterior means lie close to the OLS estimates.

```{r}
par(mfrow = c(2,3))

for (j in 1:ncol(par_data$beta)) {
  
  dens <- density(par_data$beta[,j])
  plot(dens, main = paste("Beta", j), xlab = "")
  
  # Credible interval
  ci <- quantile(par_data$beta[,j], c(0.025, 0.975))
  abline(v = ci, lty = 1)
  
  # OLS confidence interval
  residuals <- Y - X %*% ols
  sigma2_hat <- sum(residuals^2) / (nrow(X) - ncol(X))
  var_beta <- sigma2_hat * solve(t(X) %*% X)
  se <- sqrt(diag(var_beta))
  
  ci_ols <- c(ols[j] - 1.96*se[j], ols[j] + 1.96*se[j])
  abline(v = ci_ols, lty = 2)
}
```

```{r}

```
