---
title: "Bayesian Assignment 1"
author: "Aphinda Jabe, Nandi Malinga, Shantel Mogowane"
format: pdf
editor: visual
---

```{r}
rm(list = ls())
# Packages required
require ( MASS )
library(tidyverse)
```

```{r}
#| echo: false

# Load the Auto MPG dataset
auto <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data",
  header = FALSE,
  sep = "",
  col.names = c("mpg", "cylinders", "displacement", "horsepower",
                "weight", "acceleration", "model_year", "origin", "car_name"),
  na.strings = "?"
)

# Remove missing values
auto <- na.omit(auto)

# Response variable
Y <- matrix(auto$mpg, ncol = 1)

# Predictor variables
Xpreds <- auto[, c("displacement", "horsepower", "weight", "acceleration")]

# Design matrix (add intercept + standardise predictors)
X <- as.matrix(cbind(1, scale(Xpreds)))

# Sample size
n <- nrow(X)

```

# Question 1: Gibbs Sampling

## a) derivation

## b) creating function for posterior samples
```{r}
gibbs <- function(y,x,n_iter, M, beta_til, a, b){ 
  n = nrow(x)
  k = ncol(x)
  
  beta_til = as.matrix(beta_til)
  
  beta_samples = matrix(0, n_iter, k)
  var_samples = numeric(n_iter)
  
  #Starting values
  beta = rep(0, k)
  var = 1
  
  for (i in 1:n_iter){ 
    mu_beta <- solve(M + t(x)%*%x)%*%(t(x)%*%y + M%*%beta_til)
    
    # Sampling the betas
    beta <- MASS::mvrnorm(1, mu = mu_beta, Sigma = var * solve(M + t(x)%*%x))
                          
    #Sample from sigmasqrd
    A2 <- t(y)%*%y + t(beta_til) %*% M %*% beta_til - t(mu_beta) %*% (M + t(x) %*%x) %*% mu_beta
    
    shape = a + n/2
    rate = b + A2/2
    
    var = 1/rgamma(1, shape = shape, rate = rate)
    
    beta_samples[i,] = beta
    var_samples[i]  = var
  }
    return(list(beta = beta_samples, sigma2 = var_samples))
  }
```

## c) generate 50000 samples
```{r}
set.seed(123)

# Priors
k = ncol(X)
beta_til = rep(0,k)
M = diag(k)
a = 1
b = 1

par_data = gibbs(Y, X, 50000, M, beta_til, a = 1, b = 1)
```

### i. trace plots for each regression coefficient
``` {r}
samples <- par_data

ylabs <- c("intercept estimate", "beta_1 estimate", "beta_2 estimate", "beta_3 estimate", "beta_4 estimate")

# Trace plots for beta
par(mfrow = c(2, 2))  

for (j in 1:ncol(samples$beta)) {
  plot(samples$beta[, j], type = "l",
       main = paste("Trace plot for beta", j-1),
       ylab = ylabs[j],
       xlab = "Iteration")
}
```

The trace plots fluctuate around a stable mean and do not show any visible trends or drift over iterations. This suggests that the Markov chains have reached a stationary distribution. Therefore, we can conclude that the Gibbs sampler appears to have converged.

### ii. density plots

#### i. posterior means and OLS estimates
```{r}

# OLS estimates
ols <- solve(t(X) %*% X) %*% t(X) %*% Y

# Posterior means
post_means <- colMeans(par_data$beta)

par(mfrow = c(2,3))  # adjust depending on number of betas

for (j in 1:ncol(par_data$beta)) {
  
  dens <- density(par_data$beta[,j])
  plot(dens, main = paste("Beta", j-1), xlab = "")
  
  # Posterior mean
  abline(v = post_means[j], lty = 1, col = "red")
  
  # OLS estimate
  abline(v = ols[j], lty = 2, col = "blue")
}
```

The posterior densities are approximately symmetric and unimodal. The posterior means lie close to the OLS estimates.

#### ii. 95% credibility and confidence intervals
```{r}
par(mfrow = c(2,3))

# OLS confidence interval
  residuals <- Y - X %*% ols
  sigma2_hat <- sum(residuals^2) / (nrow(X) - ncol(X))
  var_beta <- sigma2_hat * solve(t(X) %*% X)
  se <- sqrt(diag(var_beta))

for (j in 1:ncol(par_data$beta)) {
  
  dens <- density(par_data$beta[,j])
  plot(dens, main = paste("Beta", j-1), xlab = "")
  
  # Credible interval
  ci <- quantile(par_data$beta[,j], c(0.025, 0.975))
  abline(v = ci, lty = 1)
  
  ci_ols <- c(ols[j] - 1.96*se[j], ols[j] + 1.96*se[j])
  abline(v = ci_ols, lty = 2)
}
```
 remember to discuss the difference!
 
#### iii. variable importance by credibility intervals

## d) informative prior
```{r}
# SAMPLING WITH INFORMATIVE PRIOR
k = ncol(X)
beta_til_inf = c(23, -3, -5, -5, 0)
Minf = 10*diag(k) # X already has intercept!
a = 1
b = 1

set.seed(123)
par_inf <- gibbs(Y, X, 50000, Minf, beta_til_inf, a, b)

# DENSITIES FROM INFORMATIVE PRIOR

# OLS estimates
ols <- solve(t(X) %*% X) %*% t(X) %*% Y

# Posterior means
post_means_inf <- colMeans(par_inf$beta)

par(mfrow = c(2,3))  # adjust depending on number of betas

for (j in 1:ncol(par_inf$beta)) {
  
  dens <- density(par_inf$beta[,j])
  plot(dens, main = paste("Beta", j-1), xlab = "")
  
  # Posterior mean
  abline(v = post_means_inf[j], lty = 1, col = "red")
  
  # OLS estimate
  abline(v = ols[j], lty = 2, col = "blue")
}
```
1. compare posteriors from vague and informative
2. how does prior affect posterior?
3. how does sample size affect posterior?
4. what happens if we only use a subset of the total data?


# Question 2
```{r}
#### Required Functions ####

library(ggplot2)
library(MASS)

#### Data Generating Functions ####

generate_lost <- function(grid_size, nsims){
  
  # Function to generate the prior distribution for the 
  # location of the lost fisherman
  # Args: 
  #       grid_size: the dimensions of the square search grid
  #       nsims: number of samples to base the prior distribution on
  
  mu_vec  <- c(grid_size/2, grid_size/2)
  sig_mat <- matrix(c(2, 1, 5, 5), 2,2)
  
  dat <- mvrnorm(nsims, mu_vec, sig_mat)
  dat <- round(abs(dat))
  
  prior <- matrix(rep(0,grid_size^2), grid_size, grid_size)
  for (i in 1:NROW(dat)){
    
    if (dat[i,1] < grid_size & dat[i,2] < grid_size){
      prior[dat[i,1], dat[i,2]] <- prior[dat[i,1], dat[i,2]] + 1
    }
    
  }
  prior <- prior/sum(prior)
  return(prior)
}

generate_fisherman <- function(grid_size){
  
  # Function to generate the true location of the lost fisherman.
  # This should not effect the search decision in any way!! It is unkown
  # to the search crew.
  # Args: 
  #       grid_size: the dimensions of the square search grid

  
  mu_vec  <- c(grid_size/2, grid_size/2)
  sig_mat <- matrix(c(2, 1, 5, 5), 2,2)
  
  location  <- round(mvrnorm(1, mu_vec, sig_mat))
  true_grid <- matrix(rep(0, grid_size^2), grid_size, grid_size)
  true_grid[location[1], location[2]] <- 1
  
  return(true_grid)
}
```


## b) Simulate bayesian search
``` {r}
# BAYESIAN SIMULATION
set.seed(123)

search_size <- 20
nsims <- 10000

theta <- generate_lost(search_size, nsims)
true_grid <- generate_fisherman(search_size)

# Detection probabilities (example: uniform)
# p <- matrix(0.6, grid_size, grid_size)
unifs <- runif(search_size^2, min = 0.6, max = 0.9)
p <- matrix(unifs, ncol = search_size)

time <- 0
found <- FALSE

theta_history <- list()
true_prob_track <- c()

true_loc <- which(true_grid == 1, arr.ind = TRUE)

while(!found){
  
  time <- time + 1
  
  # Step 1: compute detection probability
  detect_prob <- theta * p
  
  # Step 2: choose best cell
  idx <- which(detect_prob == max(detect_prob), arr.ind = TRUE)[1,]
  i <- idx[1]
  j <- idx[2]
  
  # Step 3: simulate detection
  if(true_grid[i,j] == 1){
    found <- rbinom(1,1,p[i,j]) == 1
  } else {
    found <- FALSE
  }
  
  # Record probability of true cell
  true_prob_track <- c(true_prob_track, theta[true_loc[1], true_loc[2]])

  
  # Step 4: update if not found
  if(!found){
    
    denom <- 1 - p[i,j]*theta[i,j]
    
    # Update searched cell
    theta[i,j] <- ((1 - p[i,j]) * theta[i,j]) / denom
    
    # Update all other cells
    for(a in 1:search_size){
      for(b in 1:search_size){
        if(!(a == i & b == j)){
          theta[a,b] <- theta[a,b] / denom
        }
      }
    }
    
      # Save heatmap states (first 4 steps)
  # if(time %in% c(1,5,10,20)){
  if(time %in% c(1,2,3,4)){
    theta_history[[length(theta_history)+1]] <- theta
  }
    
  }
}
```

### i. heatmaps of occurrence probability
``` {r}
#========
#  i
#=========
plot_heatmap <- function(mat, title){
  df <- reshape2::melt(mat)
  ggplot(df, aes(Var1, Var2, fill=value)) +
    geom_tile() +
    scale_fill_viridis_c() +
    ggtitle(title) +
    theme_minimal()
}

plot_heatmap(theta_history[[1]], "Step 1")
plot_heatmap(theta_history[[2]], "Step 2")
plot_heatmap(theta_history[[3]], "Step 3")
plot_heatmap(theta_history[[4]], "Step 4")

# plot_heatmap(theta_history[[1]], "Step 1")
# plot_heatmap(theta_history[[2]], "Step 5")
# plot_heatmap(theta_history[[3]], "Step 10")
# plot_heatmap(theta_history[[4]], "Step 20")
```

### ii. posterior probability of occurence
``` {r}
#=====
#ii
#======

plot(true_prob_track, type="l",
     xlab="Search Step",
     ylab="Posterior Probability",
     main="Probability of True Cell Over Time")

```


## c) comparing Bayesian and Naive strategy
```{r}
# UPDATING EQUATIONS

# new theta value for current cell
theta_newi <- function(pi, theta_oldi) {
  theta_newi <- ((1 - pi)*theta_oldi)/(1 - pi*theta_oldi)
  return(theta_newi)
}

# new theta value for all other cells
theta_newj <- function(pi, theta_oldi, theta_oldj) {
  theta_newj <- theta_oldj/(1 - pi*theta_oldi)
  return(theta_newj)
}
```

### i. 100 simulations each
```{r}
iter <- 100

# BAYESIAN 100 SIMULATIONS
set.seed(123)
bayes_results <- numeric(iter)

for (i in 1:iter) {
  
  # prior probability distributions (before search begins)
  theta <- generate_lost(search_size, nsims)
  
  # true location of fisherman 
  Y <- generate_fisherman(search_size)
  
  # probabilities of detection given that he's in the cell
  # p <- detect_pr (NANDI CODE!, p initialised earlier)
  
  # true detection (=1 if fisherman is detected, always 0 when Y is 0)
  # Z <- matrix(rbinom(search_size^2, size = 1, prob = Y*p), nrow = search_size, ncol = search_size)
  
  
  # cell with highest probability of fisherman (before search)
  index <- which(theta == max(theta), arr.ind = TRUE)[1, ]
  row <- index[1]
  col <- index[2]
  
  # indexes searched
  # searched <- index
  
  # iterators
  steps <- 0 # same as hours elapsed
  

  while (Y[row, col] != 1) {
    p_i <- p[row, col]
    old_i <- theta[row, col]
    
    # Z (was it detected? = 1 if yes (Y = 1))
    Z_i <- rbinom(1, size = 1, prob = Y[row, col]*p_i)
    
    current <- matrix(TRUE, nrow = search_size, ncol = search_size)
    current[row, col] <- FALSE
    
    theta[current] <- theta_newj(p_i, old_i, theta[current])
    
    theta[row, col] <- theta_newi(p_i, old_i)
    
    steps <- steps + 1
    #hours <- hours + 1
    #searched <- rbind(searched, index)
    
    index <- which(theta == max(theta), arr.ind = TRUE)[1, ]
    row <- index[1]
    col <- index[2]
  }
  
  bayes_results[i] <- steps
}

cat("Average number of hours:", mean(bayes_results), "\n")
```


```{r}
# NAIVE 100 SIMULATIONS

set.seed(123)
naive_results <- numeric(iter)

for (i in 1:iter) {
  
  # prior probability distributions (before search begins)
  theta <- generate_lost(search_size, nsims)
  
  # true location of fisherman 
  Y <- generate_fisherman(search_size)
  
  # probabilities of detection given that he's in the cell
  # p <- detect_pr
  
  # true detection (=1 if fisherman is detected, always 0 when Y is 0)
  # Z <- matrix(rbinom(search_size^2, size = 1, prob = Y*p), nrow = search_size, ncol = search_size)
  
  # cell with highest probability of fisherman (before search)
  # row_n <- sample(1:search_size, size = 1)
  # col_n <- sample(1:search_size, size = 1)
  # index_n <- c(row_n, col_n)
  
  cell_order <- sample(search_size^2)
  
  # iterators
  steps <- 0
  step_num <- 1
  
  # indexes
  ind_linear <- cell_order[step_num]
  index_n <- arrayInd(ind_linear, .dim = c(search_size, search_size))
  row_n <- index_n[1]
  col_n <- index_n[2]
  
  
  while (Y[row_n, col_n] != 1) {
    p_i <- p[row_n, col_n]
    old_i <- theta[row_n, col_n]
    
    # Z (was it detected? = 1 if yes (Y = 1))
    Z_i <- rbinom(1, size = 1, prob = Y[row_n, col_n]*p_i)
    
    current <- matrix(TRUE, nrow = search_size, ncol = search_size)
    current[row_n, col_n] <- FALSE
    
    theta[current] <- theta_newj(p_i, old_i, theta[current])
    theta[row_n, col_n] <- theta_newi(p_i, old_i)
    
    steps <- steps + 1
    step_num <- step_num + 1
    
    # hours <- hours + 1
    # searched <- rbind(searched, index_n)
    
    # next cell
    ind_linear <- cell_order[step_num]
    row_n <- ((ind_linear - 1) %% search_size) + 1
    col_n <- ((ind_linear - 1) %/% search_size) + 1
    
    # row_n <- sample(1:search_size, size = 1)
    # col_n <- sample(1:search_size, size = 1)
    # index_n <- c(row_n, col_n)
  }
  
  naive_results[i] <- steps
}

cat("Average number of hours:", mean(naive_results), "\n")
```

# Question 3

## a) derivation for conditional posterior distributions

## b) Code for samples
```{r}
set.seed(123)
n1 <-350 # Number of 'standard' observations
n2 <-n- n1 # Number of outliers
outlier.idx <-(n1 + 1):n # Index set I2
# Inflate the response for the outlier observations
Y.contaminated <-Y
Y.contaminated[outlier.idx] <-Y[outlier.idx] +
  rnorm(n2, mean = 0, sd = 15)

X1 <- X[1:n1,]
X2 <- X[(n1+1):n,]
y1 <- Y.contaminated[1:n1,]
y2 <- Y.contaminated[outlier.idx]

a1 <- 0.1
b1 <- 0.1
a2 <- 0.1
b2 <- 0.1
T_0 <- diag(0.001, ncol(X1))
```

## c) posterior samples

``` {r}
iterations <- 70000
betas <- matrix(0, iterations, ncol(X1))
tau1s <- tau2s <- numeric(iterations)

tau1 <- 1
tau2 <- 1

for (i in 1:iterations) {
  
  post_variance <- solve(tau1 * crossprod(X1) + tau2 * crossprod(X2) + solve(T_0))
  mu <- post_variance %*% (tau1 * t(X1) %*% y1 + tau2 * t(X2) %*% y2)
  beta <- mvrnorm(1, mu, post_variance)
  
  rss1 <- sum((y1 - X1 %*% beta)^2)
  tau1 <- rgamma(1, a1 + n1/2, b1 + 0.5*rss1)
  
  rss2 <- sum((y2 - X2 %*% beta)^2)
  tau2 <- rgamma(1, a2 + n2/2, b2 + 0.5*rss2)
  
  betas[i,] <- beta
  tau1s[i] <- tau1
  tau2s[i] <- tau2
  
}
```

### i. trace plots
``` {r}
betas <- betas[(0.2 * iterations + 1):iterations,]
tau1s <- tau1s[(0.2 * iterations + 1):iterations]
tau2s <- tau2s[(0.2 * iterations + 1):iterations]

par(mfrow = c(1, 3))
plot(betas[,1], type = "l")
plot(tau1s, type = "l")
plot(tau2s, type = "l")
```

