---
title: "Bayesian Assignment 1"
author: "Aphinda Jabe, Nandi Malinga, Shantel Mogowane"
format: pdf
editor: visual
---

```{r}
rm(list = ls())
# Packages required
require ( MASS )
library(tidyverse)
```

```{r}
#| echo: false

# Load the Auto MPG dataset
auto <- read.csv(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/auto-mpg/auto-mpg.data",
  header = FALSE,
  sep = "",
  col.names = c("mpg", "cylinders", "displacement", "horsepower",
                "weight", "acceleration", "model_year", "origin", "car_name"),
  na.strings = "?"
)

# Remove missing values
auto <- na.omit(auto)

# Response variable
Y <- matrix(auto$mpg, ncol = 1)

# Predictor variables
Xpreds <- auto[, c("displacement", "horsepower", "weight", "acceleration")]

# Design matrix (add intercept + standardise predictors)
X <- as.matrix(cbind(1, scale(Xpreds)))

# Sample size
n <- nrow(X)

```

# Question 1: Gibbs Sampling

## a) derivation

## b) creating function for posterior samples

```{r}
gibbs <- function(y,x,n_iter, M, beta_til, a, b){ 
  n = nrow(x)
  k = ncol(x)
  
  beta_til = as.matrix(beta_til)
  
  beta_samples = matrix(0, n_iter, k)
  var_samples = numeric(n_iter)
  
  #Starting values
  beta = rep(0, k)
  var = 1
  
  for (i in 1:n_iter){ 
    mu_beta <- solve(M + t(x)%*%x)%*%(t(x)%*%y + M%*%beta_til)
    
    # Sampling the betas
    beta <- MASS::mvrnorm(1, mu = mu_beta, Sigma = var * solve(M + t(x)%*%x))
                          
    #Sample from sigmasqrd
    A2 <- t(y)%*%y + t(beta_til) %*% M %*% beta_til - t(mu_beta) %*% (M + t(x) %*%x) %*% mu_beta
    
    shape = a + n/2
    rate = b + A2/2
    
    var = 1/rgamma(1, shape = shape, rate = rate)
    
    beta_samples[i,] = beta
    var_samples[i]  = var
  }
    return(list(beta = beta_samples, sigma2 = var_samples))
  }
```

## c) generate 50000 samples

```{r}
set.seed(123)

# Priors
k = ncol(X)
beta_til = rep(0,k)
M = diag(k)
a = 1
b = 1

par_data = gibbs(Y, X, 50000, M, beta_til, a = 1, b = 1)
```

### i. trace plots for each regression coefficient

```{r}
samples <- par_data

ylabs <- c("intercept estimate", "beta_1 estimate", "beta_2 estimate", "beta_3 estimate", "beta_4 estimate")

# Trace plots for beta
pdf("trace_plots.pdf", width = 10, height = 6)

par(mfrow = c(2, 3))  

for (j in 1:ncol(samples$beta)) {
  plot(samples$beta[, j], type = "l",
       main = paste("Trace plot for beta", j-1),
       ylab = ylabs[j],
       xlab = "Iteration")
}

dev.off()
```

The trace plots fluctuate around a stable mean and do not show any visible trends or drift over iterations. This suggests that the Markov chains have reached a stationary distribution. Therefore, we can conclude that the Gibbs sampler appears to have converged.

### ii. density plots

#### i. posterior means and OLS estimates

```{r}

# OLS estimates
ols <- solve(t(X) %*% X) %*% t(X) %*% Y

# Posterior means
post_means <- colMeans(par_data$beta)

pdf("beta_densities.pdf", width = 10, height = 6)

par(mfrow = c(2,3))  # adjust depending on number of betas

for (j in 1:ncol(par_data$beta)) {
  
  dens <- density(par_data$beta[,j])
  plot(dens, main = paste("Beta", j-1), xlab = "")
  
  # Posterior mean
  abline(v = post_means[j], lty = 1, col = "red")
  
  # OLS estimate
  abline(v = ols[j], lty = 2, col = "purple")
  
  legend("bottomright",
       legend = c("posterior", "OLS"),
       col = c("red", "purple"),
       lty = c(1, 2), cex = 0.85, bty = "n")
}

dev.off()
```

The plots above indicate that the posterior densities are approximately symmetric and unimodal. The posterior means closely match the OLS estimates, indicating that the likelihood is highly informative relative to the prior. The prior therefore has little influence on the posterior.

#### ii. 95% credibility and confidence intervals

```{r}
par(mfrow = c(2,3))

# OLS confidence interval
  residuals <- Y - X %*% ols
  sigma2_hat <- sum(residuals^2) / (nrow(X) - ncol(X))
  var_beta <- sigma2_hat * solve(t(X) %*% X)
  se <- sqrt(diag(var_beta))

for (j in 1:ncol(par_data$beta)) {
  
  dens <- density(par_data$beta[,j])
  plot(dens, main = paste("Beta", j-1), xlab = "")
  
  # Credible interval
  ci <- quantile(par_data$beta[,j], c(0.025, 0.975))
  abline(v = ci, lty = 1, col = "hotpink")
  
  ci_ols <- c(ols[j] - 1.96*se[j], ols[j] + 1.96*se[j])
  abline(v = ci_ols, lty = 2, col = "blue4")
  
  legend("bottomright",
       legend = c("credibility", "confidence"),
       col = c("hotpink", "blue4"),
       lty = c(1, 2), cex = 0.85, bty = "n")
}
```

remember to discuss the difference!

#### iii. variable importance by credibility intervals

## d) informative prior

```{r}
# SAMPLING WITH INFORMATIVE PRIOR
k = ncol(X)
beta_til_inf = c(23, -3, -5, -5, 0)
Minf = 10*diag(k) # X already has intercept!
a = 1
b = 1

set.seed(123)
par_inf <- gibbs(Y, X, 50000, Minf, beta_til_inf, a, b)

# DENSITIES FROM INFORMATIVE PRIOR

# OLS estimates
ols <- solve(t(X) %*% X) %*% t(X) %*% Y

# Posterior means
post_means_inf <- colMeans(par_inf$beta)

par(mfrow = c(2,3))  # adjust depending on number of betas

for (j in 1:ncol(par_inf$beta)) {
  
  dens <- density(par_inf$beta[,j])
  plot(dens, main = paste("Beta", j-1), xlab = "")
  
  # Posterior mean
  abline(v = post_means_inf[j], lty = 1, col = "red")
  
  # OLS estimate
  abline(v = ols[j], lty = 2, col = "blue")
  
  legend("bottomright",
       legend = c("posterior", "OLS"),
       col = c("red", "blue"),
       lty = c(1, 2), cex = 0.85, bty = "n")
}
```

1.  compare posteriors from vague and informative
2.  how does prior affect posterior?
3.  how does sample size affect posterior?
4.  what happens if we only use a subset of the total data?

# Question 2

```{r}
#### Required Functions ####

library(ggplot2)
library(MASS)

#### Data Generating Functions ####

generate_lost <- function(grid_size, nsims){
  
  # Function to generate the prior distribution for the 
  # location of the lost fisherman
  # Args: 
  #       grid_size: the dimensions of the square search grid
  #       nsims: number of samples to base the prior distribution on
  
  mu_vec  <- c(grid_size/2, grid_size/2)
  sig_mat <- matrix(c(2, 1, 5, 5), 2,2)
  
  dat <- mvrnorm(nsims, mu_vec, sig_mat)
  dat <- round(abs(dat))
  
  prior <- matrix(rep(0,grid_size^2), grid_size, grid_size)
  for (i in 1:NROW(dat)){
    
    if (dat[i,1] < grid_size & dat[i,2] < grid_size){
      prior[dat[i,1], dat[i,2]] <- prior[dat[i,1], dat[i,2]] + 1
    }
    
  }
  prior <- prior/sum(prior)
  return(prior)
}

generate_fisherman <- function(grid_size){
  
  # Function to generate the true location of the lost fisherman.
  # This should not effect the search decision in any way!! It is unkown
  # to the search crew.
  # Args: 
  #       grid_size: the dimensions of the square search grid

  
  mu_vec  <- c(grid_size/2, grid_size/2)
  sig_mat <- matrix(c(2, 1, 5, 5), 2,2)
  
  location  <- round(mvrnorm(1, mu_vec, sig_mat))
  true_grid <- matrix(rep(0, grid_size^2), grid_size, grid_size)
  true_grid[location[1], location[2]] <- 1
  
  return(true_grid)
}
```

## b) Simulate bayesian search

```{r}
# BAYESIAN SIMULATION
set.seed(123)

search_size <- 20
nsims <- 10000

theta <- generate_lost(search_size, nsims)
true_grid <- generate_fisherman(search_size)

# Detection probabilities (example: uniform)
# p <- matrix(0.6, grid_size, grid_size)
unifs <- runif(search_size^2, min = 0.6, max = 0.9)
p <- matrix(unifs, ncol = search_size)

time <- 0
found <- FALSE

theta_history <- list()
true_prob_track <- c()

true_loc <- which(true_grid == 1, arr.ind = TRUE)

while(!found){
  
  time <- time + 1
  
  # Step 1: compute detection probability
  detect_prob <- theta * p
  
  # Step 2: choose best cell
  idx <- which(detect_prob == max(detect_prob), arr.ind = TRUE)[1,]
  i <- idx[1]
  j <- idx[2]
  
  # Step 3: simulate detection
  if(true_grid[i,j] == 1){
    found <- rbinom(1,1,p[i,j]) == 1
  } else {
    found <- FALSE
  }
  
  # Record probability of true cell
  true_prob_track <- c(true_prob_track, theta[true_loc[1], true_loc[2]])

  
  # Step 4: update if not found
  if(!found){
    
    denom <- 1 - p[i,j]*theta[i,j]
    
    # Update searched cell
    theta[i,j] <- ((1 - p[i,j]) * theta[i,j]) / denom
    
    # Update all other cells
    for(a in 1:search_size){
      for(b in 1:search_size){
        if(!(a == i & b == j)){
          theta[a,b] <- theta[a,b] / denom
        }
      }
    }
    
      # Save heatmap states (first 4 steps)
  # if(time %in% c(1,5,10,20)){
  if(time %in% c(1,2,3,4)){
    theta_history[[length(theta_history)+1]] <- theta
  }
    
  }
}
```

### i. heatmaps of occurrence probability

```{r}
#========
#  i
#=========
plot_heatmap <- function(mat, title){
  df <- reshape2::melt(mat)
  ggplot(df, aes(Var1, Var2, fill=value)) +
    geom_tile() +
    scale_fill_viridis_c() +
    ggtitle(title) +
    theme_minimal()
}

plot_heatmap(theta_history[[1]], "Step 1")
plot_heatmap(theta_history[[2]], "Step 2")
plot_heatmap(theta_history[[3]], "Step 3")
plot_heatmap(theta_history[[4]], "Step 4")

# plot_heatmap(theta_history[[1]], "Step 1")
# plot_heatmap(theta_history[[2]], "Step 5")
# plot_heatmap(theta_history[[3]], "Step 10")
# plot_heatmap(theta_history[[4]], "Step 20")
```

### ii. posterior probability of occurence

```{r}
#=====
#ii
#======

plot(true_prob_track, type="l",
     xlab="Search Step",
     ylab="Posterior Probability",
     main="Probability of True Cell Over Time")

```

## c) comparing Bayesian and Naive strategy

```{r}
# UPDATING EQUATIONS

# new theta value for current cell
theta_newi <- function(pi, theta_oldi) {
  theta_newi <- ((1 - pi)*theta_oldi)/(1 - pi*theta_oldi)
  return(theta_newi)
}

# new theta value for all other cells
theta_newj <- function(pi, theta_oldi, theta_oldj) {
  theta_newj <- theta_oldj/(1 - pi*theta_oldi)
  return(theta_newj)
}
```

### i. 100 simulations each

```{r}
iter <- 100

# BAYESIAN 100 SIMULATIONS
set.seed(123)
bayes_results <- numeric(iter)

for (i in 1:iter) {
  
  # prior probability distributions (before search begins)
  theta <- generate_lost(search_size, nsims)
  
  # true location of fisherman 
  Y <- generate_fisherman(search_size)
  
  # probabilities of detection given that he's in the cell
  # p <- detect_pr (NANDI CODE!, p initialised earlier)
  
  # true detection (=1 if fisherman is detected, always 0 when Y is 0)
  # Z <- matrix(rbinom(search_size^2, size = 1, prob = Y*p), nrow = search_size, ncol = search_size)
  
  
  # cell with highest probability of fisherman (before search)
  index <- which(theta == max(theta), arr.ind = TRUE)[1, ]
  row <- index[1]
  col <- index[2]
  
  # indexes searched
  # searched <- index
  
  # iterators
  steps <- 0 # same as hours elapsed
  

  while (Y[row, col] != 1) {
    p_i <- p[row, col]
    old_i <- theta[row, col]
    
    # Z (was it detected? = 1 if yes (Y = 1))
    Z_i <- rbinom(1, size = 1, prob = Y[row, col]*p_i)
    
    current <- matrix(TRUE, nrow = search_size, ncol = search_size)
    current[row, col] <- FALSE
    
    theta[current] <- theta_newj(p_i, old_i, theta[current])
    
    theta[row, col] <- theta_newi(p_i, old_i)
    
    steps <- steps + 1
    #hours <- hours + 1
    #searched <- rbind(searched, index)
    
    index <- which(theta == max(theta), arr.ind = TRUE)[1, ]
    row <- index[1]
    col <- index[2]
  }
  
  bayes_results[i] <- steps
}

cat("Average number of hours:", mean(bayes_results), "\n")
```

```{r}
# NAIVE 100 SIMULATIONS

set.seed(123)
naive_results <- numeric(iter)

for (i in 1:iter) {
  
  # prior probability distributions (before search begins)
  theta <- generate_lost(search_size, nsims)
  
  # true location of fisherman 
  Y <- generate_fisherman(search_size)
  
  # probabilities of detection given that he's in the cell
  # p <- detect_pr
  
  # true detection (=1 if fisherman is detected, always 0 when Y is 0)
  # Z <- matrix(rbinom(search_size^2, size = 1, prob = Y*p), nrow = search_size, ncol = search_size)
  
  # cell with highest probability of fisherman (before search)
  # row_n <- sample(1:search_size, size = 1)
  # col_n <- sample(1:search_size, size = 1)
  # index_n <- c(row_n, col_n)
  
  cell_order <- sample(search_size^2)
  
  # iterators
  steps <- 0
  step_num <- 1
  
  # indexes
  ind_linear <- cell_order[step_num]
  index_n <- arrayInd(ind_linear, .dim = c(search_size, search_size))
  row_n <- index_n[1]
  col_n <- index_n[2]
  
  
  while (Y[row_n, col_n] != 1) {
    p_i <- p[row_n, col_n]
    old_i <- theta[row_n, col_n]
    
    # Z (was it detected? = 1 if yes (Y = 1))
    Z_i <- rbinom(1, size = 1, prob = Y[row_n, col_n]*p_i)
    
    current <- matrix(TRUE, nrow = search_size, ncol = search_size)
    current[row_n, col_n] <- FALSE
    
    theta[current] <- theta_newj(p_i, old_i, theta[current])
    theta[row_n, col_n] <- theta_newi(p_i, old_i)
    
    steps <- steps + 1
    step_num <- step_num + 1
    
    # hours <- hours + 1
    # searched <- rbind(searched, index_n)
    
    # next cell
    ind_linear <- cell_order[step_num]
    row_n <- ((ind_linear - 1) %% search_size) + 1
    col_n <- ((ind_linear - 1) %/% search_size) + 1
    
    # row_n <- sample(1:search_size, size = 1)
    # col_n <- sample(1:search_size, size = 1)
    # index_n <- c(row_n, col_n)
  }
  
  naive_results[i] <- steps
}

cat("Average number of hours:", mean(naive_results), "\n")
```

# Question 3

## a) derivation for conditional posterior distributions

#### First derivation

$$p(\beta \mid \tau_1, \tau_2, X, Y) \propto p(Y \mid X, \beta, \tau_1, \tau_2) \, p(\beta)$$

$$p(Y \mid X, \beta, \tau_1, \tau_2) \propto \exp\left\{ -\frac{1}{2} \left[ \tau_1 (Y_1 - X_1\beta)^T (Y_1 - X_1\beta) + \tau_2 (Y_2 - X_2\beta)^T (Y_2 - X_2\beta) \right] \right\}$$

$$p(\beta \mid \tau_1, \tau_2, X, Y) \propto \exp\left\{ -\frac{1}{2} \left[ \beta^T T_0^{-1} \beta + \tau_1 (Y_1 - X_1\beta)^T (Y_1 - X_1\beta) + \tau_2 (Y_2 - X_2\beta)^T (Y_2 - X_2\beta) \right] \right\}$$

$$\propto \exp\left\{ -\frac{1}{2} \left[ \beta^T \left( \tau_1 X_1^T X_1 + \tau_2 X_2^T X_2 + T_0^{-1} \right) \beta - 2\beta^T \left( \tau_1 X_1^T Y_1 + \tau_2 X_2^T Y_2 \right) \right] \right\}$$

**The conditional posterior is:** $$\beta \mid \tau_1, \tau_2, X, Y \sim N(\mu_n, \Sigma_n)$$ where $$\Sigma_n^{-1} = \tau_1 X_1^T X_1 + \tau_2 X_2^T X_2 + T_0^{-1}$$ $$\mu_n = \Sigma_n \left( \tau_1 X_1^T Y_1 + \tau_2 X_2^T Y_2 \right)$$

#### Second derivation

$$p(\tau_1 \mid \beta, \tau_2, X, Y) \propto p(Y_1 \mid X_1, \beta, \tau_1) \, p(\tau_1)$$

$$p(\tau_1) \propto \tau_1^{a-1} \exp(-b\,\tau_1)$$

\
$$p(Y_1 \mid X_1, \beta, \tau_1) \propto \tau_1^{n_1/2} \exp\left\{ -\frac{\tau_1}{2} (Y_1 - X_1\beta)^T (Y_1 - X_1\beta) \right\}$$ $$= \tau_1^{n_1/2} \exp\left\{ -\frac{\tau_1}{2} \sum_{i=1}^{n_1} (y_{1i} - x_{1i}^T\beta)^2 \right\}$$

$$p(\tau_1 \mid \beta, \tau_2, X, Y) \propto \tau_1^{a-1} \exp(-b\,\tau_1) \times \tau_1^{n_1/2} \exp\left\{ -\frac{\tau_1}{2} \sum_{i=1}^{n_1} (y_{1i} - x_{1i}^T\beta)^2 \right\}$$ $$\propto \tau_1^{a + n_1/2 - 1} \exp\left\{ -\tau_1 \left[ b + \frac{1}{2} \sum_{i=1}^{n_1} (y_{1i} - x_{1i}^T\beta)^2 \right] \right\}$$

**The conditional posterior is:** $$\tau_1 \mid \beta, \tau_2, X, Y \sim \text{Gamma}\left( a + \frac{n_1}{2}, \; b + \frac{1}{2} \sum_{i=1}^{n_1} (y_{1i} - x_{1i}^T\beta)^2 \right)$$

#### Third derivation

$$p(\tau_2 \mid \tau_1, \beta, X, Y) \propto p(Y_2 \mid X_2, \beta, \tau_2) \, p(\tau_2 \mid \tau_1)$$

$$p(\tau_2 \mid \tau_1) \propto \tau_2^{a-1} \exp(-b\,\tau_2) \, \mathbb{I}(0 < \tau_2 < \tau_1)$$

$$p(Y_2 \mid X_2, \beta, \tau_2) \propto \tau_2^{n_2/2} \exp\left\{ -\frac{\tau_2}{2} (Y_2 - X_2\beta)^T (Y_2 - X_2\beta) \right\}$$

$$p(\tau_2 \mid \tau_1, \beta, X, Y) \propto \tau_2^{a + n_2/2 - 1} \exp\left\{ -\tau_2 \left[ b + \frac{1}{2} (Y_2 - X_2\beta)^T (Y_2 - X_2\beta) \right] \right\} \mathbb{I}(0 < \tau_2 < \tau_1)$$

**The conditional posterior is:** $$\tau_2 \mid \tau_1, \beta, X, Y \sim \left( a + \frac{n_2}{2}, \; b + \frac{1}{2} \sum_{i=1}^{n_2} (y_{2i} - x_{2i}^T\beta)^2 \right) \mathbb{I}(0 < \tau_2 < \tau_1)$$ \## b) Code for samples $$
\begin{align*}p(\beta, \tau_1, \tau_2 \mid X, Y)&\propto p(Y \mid X, \beta, \tau_1, \tau_2) \, p(\beta) \, p(\tau_1, \tau_2) \\\\\quad p(Y \mid X, \beta, \tau_1, \tau_2)&\propto \tau_1^{n_1/2} \tau_2^{(n-n_1)/2}\exp\left\{ -\frac{1}{2} \left[\tau_1 (Y_1 - X_1\beta)^T (Y_1 - X_1\beta)+ \tau_2 (Y_2 - X_2\beta)^T (Y_2 - X_2\beta)\right] \right\} \\\\p(\beta) &\propto \exp\left\{ -\frac{1}{2} \beta^T T_0^{-1} \beta \right\} \\\\p(\tau_1, \tau_2) &\propto \tau_1^{a-1} \tau_2^{a-1} \exp\left\{ -b(\tau_1 + \tau_2) \right\} \mathbb{I}(0 < \tau_2 < \tau_1) \\\\\quad p(\beta, \tau_1, \tau_2 \mid X, Y)&\propto \tau_1^{a + n_1/2 - 1} \tau_2^{a + (n-n_1)/2 - 1} \\&\quad \times \exp\left\{ -\frac{1}{2}\beta^T T_0^{-1} \beta- \frac{\tau_1}{2} (Y_1 - X_1\beta)^T (Y_1 - X_1\beta)- \frac{\tau_2}{2} (Y_2 - X_2\beta)^T (Y_2 - X_2\beta)- b(\tau_1 + \tau_2) \right\} \\&\quad \times \mathbb{I}(0 < \tau_2 < \tau_1)\end{align*}
$$

```{r}

# Contaminate the data with outliers
set.seed(12)
n1 <-350 # Number of 'standard' observations
n2 <-n- n1 # Number of outliers
outlier.idx <-(n1 + 1):n # Index set I2
# Inflate the response for the outlier observations
Y.contaminated <-Y
Y.contaminated[outlier.idx] <-Y[outlier.idx] +
  rnorm(n2, mean = 0, sd = 15)

X1 <- X[1:n1,]
X2 <- X[(n1+1):n,]
y1 <- Y.contaminated[1:n1,]
y2 <- Y.contaminated[outlier.idx]

a1 <- 0.1
b1 <- 0.1
a2 <- 0.1
b2 <- 0.1
T_0 <- diag(0.001, ncol(X1))

iterations <- 70000
betas <- matrix(0, iterations, ncol(X1))
tau1s <- tau2s <- numeric(iterations)

tau1 <- 1
tau2 <- 0.5

for (i in 1:iterations) {
  
  post_variance <- solve(tau1 * crossprod(X1) + tau2 * crossprod(X2) + solve(T_0))
  mu <- post_variance %*% (tau1 * t(X1) %*% y1 + tau2 * t(X2) %*% y2)
  beta <- mvrnorm(1, mu, post_variance)
  
  rss1 <- sum((y1 - X1 %*% beta)^2)
  tau1 <- rgamma(1, a1 + n1/2, b1 + 0.5*rss1)
  
  rss2 <- sum((y2 - X2 %*% beta)^2)
  repeat {tau2_temp <- rgamma(1, shape = a2 + n2/2, rate = b2 + 0.5*rss2)
  if (tau2_temp <- tau1) break}
  tau2 <- tau2_temp
  
  betas[i,] <- beta
  tau1s[i] <- tau1
  tau2s[i] <- tau2
  
}

betas <- betas[(0.2 * iterations + 1):iterations,]
tau1s <- tau1s[(0.2 * iterations + 1):iterations]
tau2s <- tau2s[(0.2 * iterations + 1):iterations]

```

## c) posterior samples

### i. trace plots

```{r, echo = FALSE}
#| fig-cap: "Trace plots for beta, tau_1, tau_2 for assessment of convergence"
par(mfrow = c(1, 3))
plot(betas[,1], type = "l")
plot(tau1s, type = "l")
plot(tau2s, type = "l")
```
